---
title: "Modelling"
output: html_notebook
---

```{r, message = FALSE}
source("functions.R")
library(tidyverse)
library(reticulate)
```

Read imputed data
```{r}
load("../data/imputed.Rda")
```

Note that some values are higher than 1, which is not possible. We have not been able to determine why Grech et al determined some values to be higher than 1, but here we will set them equal to one.

```{r}
imputed$mean.phylop[imputed$mean.phylop > 1] <- 1
```


## Train-test split
```{r}
split <- train_test_split(imputed, 0.2)
train <- split$train %>% select(-Systematic_ID)
test <- split$test %>% select(-Systematic_ID)
```

# Train models
```{r}
pls_mod <- pls::plsr(mean.phylop ~ ., data = train,
                     scale = TRUE, center = TRUE, validation = "CV")

pcr_mod <- pls::pcr(mean.phylop ~ ., data = train,
                     scale = TRUE, center = TRUE, validation = "CV")

X_cols <- train %>% select(-mean.phylop)
rf_mod <- randomForest::randomForest(x = X_cols, y = train$mean.phylop,
                     ntree = 500, mtry = floor(ncol(X_cols)/3))

```

# Evaluate models
```{r}
eval <- evaluate_mvr(pls_mod, test, "mean.phylop", "PLS") %>%
  add_row(evaluate_mvr(pcr_mod, test, "mean.phylop", "PCR")) %>%
  add_row(evaluate_rf(rf_mod, test, "mean.phylop", "RF")) %>%
  write_csv(., "../data/model_performance.csv")
```

## VIP scores (PLS)
```{r}
var_imp<- VIP(pls_mod, eval$ncomps[eval$model == "PLS"]) %>%
  data.frame(t(.)) %>% select(1) %>% rownames_to_column()
colnames(var_imp) <- c("var", "VIP_score")
```

Assign broad variable categories

```{r}
process <- var_imp[grep("Process", var_imp$var),] %>% add_column(var_group = "Process")
funct <- var_imp[grep("Function", var_imp$var),] %>% add_column(var_group = "Function")
component <- var_imp[grep("Component", var_imp$var),] %>% add_column(var_group = "Component")
size <- var_imp %>% filter(var == "genelength" | var == "Mass..kDa." | var == "Residues") %>% add_column(var_group = "Size")
importance <- var_imp %>% filter(var == "essential1" | var == "solid.med.fitness") %>% add_column(var_group = "Functional importance")
expression <- var_imp %>% filter(var == "log.phase.RPKM" | var == "sum.protein.cpc" | var == "sum.mRNA.cpc")  %>% add_column(var_group = "Expression")
charge <- var_imp %>% filter(var == "pI" | var == "Charge") %>% add_column(var_group = "Charge")
amino_acids <- var_imp %>% filter(nchar(var) == 1) %>% add_column(var_group = "Amino acid content")
network <- var_imp %>% filter(str_detect(var, "phys_") | str_detect(var, "funct_"))  %>% add_column(var_group = "Network centrality")
location <- var_imp %>% filter(var == "chrI1" | var == "chrII1" | var == "chrIII1" | var == "genomic_location" | var == "Rel_telomere" | var == "Rel_centromere") %>% add_column(var_group = "Genomic location")
codon_bias <- var_imp %>% filter(var == "CAI") %>% add_column(var_group = "Codon bias")
introns <- var_imp %>% filter(var == "NumberIntrons" | var == "AvergaeIntLength") %>% add_column(var_group = "Introns")
```

Combine as single df
```{r}
variable_importance <- rbind(process, funct, component, size, importance, expression, charge, amino_acids, network, location, codon_bias, introns)

variable_importance$var <- str_remove_all(variable_importance$var, "1")
variable_importance$var <- str_remove_all(variable_importance$var, "`")
  
```


## % variance explained per variable (PCR)

Export output to Python (because string handling in R is beyond useless)
```{r}
pcr_summary <- capture.output(summary(pcr_mod))
py_summary <- r_to_py(pcr_summary)
```

```{python}
lines = r.py_summary
res = [line.split() for line in lines if "mean.phylop" in line.split()]
res = [item for sublist in res for item in sublist if "mean.phylop" not in item]
``` 

Back to R
```{r}
cumul_var_by_comp <- as.numeric(py$res)
```

Extract variance per component as well as projections
```{r, results = 'hide'}
variance_by_component <- get_variance_per_component(pcr_mod, opt_comps = eval$ncomps[eval$model == "PCR"], cumul_var_by_comp)
projections <- get_projections(pcr_mod, opt_comps = eval$ncomps[eval$model == "PCR"]) 
write.csv(projections, "../data/pcr_projections.csv")
```

Calculate variance explained by each variable
```{r}
variance_explained_per_variable = c()
for (i in 1:dim(projections)[1]) {
  variance_explained_per_variable[i] = sum(variance_by_component * projections[i, ])
}
```

Make a dataframe
```{r}
variances_pcr <- data.frame(variance_explained_per_variable) %>%
  add_column(var = rownames(projections))

colnames(variances_pcr)[1] <- ("pcr_variance_expl")

variances_pcr$var <- str_remove_all(variances_pcr$var, "1")
variances_pcr$var <- str_remove_all(variances_pcr$var, "`")
```

Combine with PLS data

```{r}
variable_importance <- variable_importance %>% merge(., variances_pcr, by = "var") 
```



## Partial correlations

Export variable groups
```{r}
variable_importance %>% select(var, var_group) %>% write.csv("../data/variable_groups.csv")
```

To obtain partial correlations, run partial_correlation.py:

python partial_correlation.py \
  --data "../data/imputed.csv" \
  --variable_groups "../data/variable_groups.csv" \
  --output "../data/partial_correlations.csv"
  

```{r}
partial_correlations <- read.csv("../data/partial_correlations.csv") %>% select(r, p.val, var, group, p.adj)
```

```{r}
partial_correlations %>% merge(., variable_importance, by = "var") %>% write.csv("../data/variable_importance.csv")
```




